---
title: 'Session 3: Reading in data and visualization'
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The first session will focus on reading in the data after `steinbock` 
segmentation (see Session 2).

# Read in the data 

We use the [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package to
read in single-cell data extracted using the `steinbock` framework. During image
processing you will also obtain multi-channel images and segmentation masks that
can be read into `R` using the
[cytomapper](https://github.com/BodenmillerGroup/cytomapper) package.

```{r, message=FALSE}
library(imcRtools)
library(cytomapper)
```

## Read in single-cell information

For single-cell data analysis in `R` the
framework. It allows standardized access to (i) expression data, (ii) cellular
metadata (e.g. cell type), (iii) feature metadata (e.g. marker name) and (iv)
experiment-wide metadata. For an in-depth introduction to the `SingleCellExperiment`
container, please refer to the [SingleCellExperiment class](https://bioconductor.org/books/3.14/OSCA.intro/the-singlecellexperiment-class.html).

The [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)
class is an extension of the `SingleCellExperiment` class. It
was developed to store spatial data in addition to single-cell data and an
extended introduction is accessible
[here](https://bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html).

To read in single-cell data generated by the `steinbock` framework the
`imcRtools` package provides the `read_steinbock` function. By default, the data
is read into a `SpatialExperiment` object; however, data can be read in as a
`SingleCellExperiment` object by setting `return_as = "sce"`. All functions
presented in this tutorial are applicable to both data containers.

### steinbock generated data

The downloaded example data (see [data](https://github.com/BodenmillerGroup/ImagingWorkshop2023/tree/main/data)) processed with the
[steinbock](https://github.com/BodenmillerGroup/steinbock) framework can be read
in with the `read_steinbock` function provided by `imcRtools`. For more
information, please refer to `?read_steinbock`.

```{r read-steinbock}
spe <- read_steinbock("data/steinbock/")
spe
```

By default, single-cell data is read in as `SpatialExperiment` object. 
The summarized pixel intensities per channel and cell (here mean intensity) are
stored in the `counts` slot. Columns represent cells and rows represent channels.

```{r counts}
counts(spe)[1:5,1:5]
```

Metadata associated to individual cells are stored in the `colData` slot. After
initial image processing, these metadata include the numeric identifier
(`ObjectNumber`), the area, and morphological features of each cell. In
addition, `sample_id` stores the image name from which each cell was extracted
and the width and height of the corresponding images are stored.

```{r colData}
head(colData(spe))
```

The main difference between the `SpatialExperiment` and the
`SingleCellExperiment` data container in the current setting is the way spatial
locations of all cells are stored. For the `SingleCellExperiment` container, the
locations are stored in the `colData` slot while the `SpatialExperiment`
container stores them in the `spatialCoords` slot:

```{r spatialCoords}
head(spatialCoords(spe))
```

The _spatial cell graphs_ generated by `steinbock` are read into a `colPair`
slot of the `SpatialExperiment` (or `SingleCellExperiment`) object. Cell-cell
interactions (cells in close spatial proximity) are represented as "edge list"
(stored as `SelfHits` object). Here, the left side represents the column indices
of the "from" cells and the right side represents the column indices of the "to"
cells. We will later see how to visualize the _spatial cell graphs_.

```{r colPair}
colPair(spe, "neighborhood")
```

Finally, metadata regarding the channels are stored in the `rowData` slot. This
information is extracted from the `panel.csv` file. Channels are ordered by
isotope mass and therefore match the channel order of the multi-channel images.

```{r rowData}
head(rowData(spe))
```

### Reading custom files

When not using `steinbock`, the single-cell
information has to be read in from custom files. We now demonstrate how
to generate a `SpatialExperiment` object from single-cell data contained
in individual files. As an example, we use files generated by `steinbock`.

First we will read in the single-cell features stored in a CSV file:

```{r read-csv, message=FALSE}
library(readr)

cur_intensities <- read_csv("data/steinbock/intensities/Patient1_001.csv")
cur_regionprobs <- read_csv("data/steinbock/regionprops/Patient1_001.csv")

dim(cur_intensities)
dim(cur_regionprobs)

colnames(cur_intensities)
colnames(cur_regionprobs)
```

These files contain single-cell features including the cell
identifier (`Object`), morphological 
features, the cells' locations (`centroid-0` and `centroid-1`) and the 
mean pixel intensity per cell and per channel (`cur_intensities`).

Now we will extract the relevant entries from the files.

```{r split-features}
counts <- cur_intensities[,-1]

meta <- cur_regionprobs[,c("Object", "area", "axis_major_length", "axis_minor_length")]

coords <- cur_regionprobs[,c("centroid-1", "centroid-0")]
```

From these features we can now construct the `SpatialExperiment` object.

```{r construct-spe, message=FALSE}
library(SpatialExperiment)
spe2 <- SpatialExperiment(assays = list(counts = t(counts)),
                          colData = meta, 
                          sample_id = "Patient1_001",
                          spatialCoords = as.matrix(coords))
```

Next, we can store the spatial cell graph generated by `steinbock` in the
`colPairs` slot of the object. Spatial cell graphs are usually stored as edge
list in form of a CSV file. The `colPairs` slot requires a `SelfHits` entry
storing an edge list where numeric entries represent the index of the `from` and
`to` cell in the `SpatialExperiment` object. To generate such an edge list, we
need to match the cell IDs contained in the CSV against the cell IDs in the
`SpatialExperiment` object.

```{r create-colPairs, message=FALSE}
cur_pairs <- read_csv("data/steinbock/neighbors/Patient1_001.csv")

edgelist <- SelfHits(from = match(cur_pairs$Object, spe2$Object),
                     to = match(cur_pairs$Neighbor, spe2$Object),
                     nnode = ncol(spe2))

colPair(spe2, "neighborhood") <- edgelist
```

For further downstream analysis, we will use the `steinbock` results as read in above.

## Single-cell processing {#cell-processing}

After reading in the single-cell data, few further processing steps need to be
taken.

**Add additional metadata**

We can set the `colnames` of the object to generate unique identifiers per cell:

```{r set-colnames}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
```

It is also often the case that sample-specific metadata are available externally.
For the current data, we need to link the cancer type (also referred to as "Indication")
to each sample. This metadata is available as external excel file:

```{r add-indication}
library(openxlsx)
library(stringr)
meta <- read.xlsx("data/steinbock/sample_metadata.xlsx")

spe$patient_id <- as.vector(str_extract_all(spe$sample_id, "Patient[1-4]", simplify = TRUE))
spe$ROI <- as.vector(str_extract_all(spe$sample_id, "00[1-8]", simplify = TRUE))
spe$indication <- meta$Indication[match(spe$patient_id, meta$Sample.ID)]

unique(spe$indication)
```

The selected patients were diagnosed with different cancer types:

* SCCHN - head and neck cancer  
* BCC - breast cancer  
* NSCLC - lung cancer  
* CRC - colorectal cancer

**Transform counts**

The distribution of expression counts across cells is often observed to be
skewed towards the right side meaning lots of cells display low counts and few
cells have high counts. To avoid analysis biases from these high-expressing
cells, the expression counts are commonly transformed or clipped.

Here, we perform counts transformation using an inverse hyperbolic sine
function. This transformation is commonly applied to [flow cytometry
data](https://support.cytobank.org/hc/en-us/articles/206148057-About-the-Arcsinh-transform).
The `cofactor` here defines the expression range on which no scaling is
performed. While the `cofactor` for CyTOF data is often set to `5`, IMC data
usually display much lower counts. We therefore apply a `cofactor` of `1`.

However, other transformations such as `log(counts(spe) + 0.01)` should be
tested when analysing IMC data.

```{r transform-counts, message=FALSE}
library(dittoSeq)
dittoRidgePlot(spe, var = "CD3", group.by = "patient_id", assay = "counts") +
    ggtitle("CD3 - before transformation")
assay(spe, "exprs") <- asinh(counts(spe)/1)
dittoRidgePlot(spe, var = "CD3", group.by = "patient_id", assay = "exprs") +
    ggtitle("CD3 - after transformation")
```

**Define interesting channels**

For downstream analysis such as visualization, dimensionality reduction and
clustering, only a subset of markers should be used. As convenience, we can
store an additional entry in the `rowData` slot that specifies the markers of
interest. Here, we deselect the nuclear markers, which were primarily used for
cell segmentation, and keep all other biological targets.

```{r select-features}
rowData(spe)$use_channel <- !grepl("DNA|Histone", rownames(spe))
```

**Define color schemes**

We will define color schemes for different metadata entries of the data and
conveniently store them in the `metadata` slot of the `SpatialExperiment` which
will be helpful for downstream data visualizations. We will use colors from the
`RColorBrewer` and `dittoSeq` package but any other coloring package will
suffice.

```{r define-colors, message=FALSE}
library(RColorBrewer)
color_vectors <- list()

ROI <- setNames(brewer.pal(length(unique(spe$ROI)), name = "BrBG"), 
                unique(spe$ROI))
patient_id <- setNames(brewer.pal(length(unique(spe$patient_id)), name = "Set1"), 
                unique(spe$patient_id))
sample_id <- setNames(c(brewer.pal(6, "YlOrRd")[3:5],
                        brewer.pal(6, "PuBu")[3:6],
                        brewer.pal(6, "YlGn")[3:5],
                        brewer.pal(6, "BuPu")[3:6]),
                unique(spe$sample_id))
indication <- setNames(brewer.pal(length(unique(spe$indication)), name = "Set2"), 
                unique(spe$indication))

color_vectors$ROI <- ROI
color_vectors$patient_id <- patient_id
color_vectors$sample_id <- sample_id
color_vectors$indication <- indication

metadata(spe)$color_vectors <- color_vectors
```

## Read in images {#read-images}

The `cytomapper` package allows multi-channel image handling and visualization
within the Bioconductor framework. The most common data format for multi-channel
images or segmentation masks is the TIFF file format, which is used by
`steinbock`.

Here, we will read in multi-channel images and segmentation masks into a
[CytoImageList](https://www.bioconductor.org/packages/release/bioc/vignettes/cytomapper/inst/doc/cytomapper.html#5_The_CytoImageList_object)
data container. It allows storing multiple multi-channel images and requires
matched channels across all images within the object.

The `loadImages` function is used to read in processed multi-channel images and
their corresponding segmentation masks. Of note, the multi-channel images
generated by `steinbock` are saved as 32-bit images while the segmentation masks
are saved as 16-bit images. To correctly scale pixel values of the segmentation
masks when reading them in set `as.is = TRUE`.

```{r read-images}
images <- loadImages("data/steinbock/img/")
masks <- loadImages("data/steinbock/masks/", as.is = TRUE)
```

In the case of multi-channel images, it is beneficial to set the `channelNames`
for easy visualization. Using the `steinbock` framework, the channel order of
the single-cell data matches the channel order of the multi-channel images.
However, it is recommended to make sure that the channel order is identical
between the single-cell data and the images.

```{r set-channelNames}
channelNames(images) <- rownames(spe)
images
```

For image and mask visualization we will need to
add additional metadata to the `elementMetadata` slot of the `CytoImageList`
objects. This slot is easily accessible using the `mcols` function.

Here, we will save the matched `sample_id`, `patient_id` and `indication`
information within the `elementMetadata` slot of the multi-channel images and
segmentation masks objects. It is crucial that the order of the images in 
both `CytoImageList` objects is the same.

```{r add-metadata}
all.equal(names(images), names(masks))
patient_id <- str_extract_all(names(images), "Patient[1-4]", simplify = TRUE)
indication <- meta$Indication[match(patient_id, meta$Sample.ID)] 

mcols(images) <- mcols(masks) <- DataFrame(sample_id = names(images),
                                           patient_id = patient_id,
                                           indication = indication)
```

## Generate single-cell data from images

An alternative way of generating a `SingleCellExperiment` object directly 
from the multi-channel images and segmentation masks is supported by the 
[measureObjects](https://bodenmillergroup.github.io/cytomapper/reference/measureObjects.html)
function of the `cytomapper` package. For each cell present in the `masks`
object, the function computes the mean pixel intensity per channel as well as
morphological features (area, radius, major axis length, eccentricity) and the 
location of cells:

```{r measureObjects, message=FALSE}
cytomapper_sce <- measureObjects(masks, image = images, img_id = "sample_id")

cytomapper_sce
```

# Single-cell visualization

The following section focuses on visualizing the single-cell data contained
in the `SpatialExperiment` object. The main R/Bioconductor packages
to support visualization are `r BiocStyle::Biocpkg("dittoSeq")` and
`r BiocStyle::Biocpkg("imcRtools")`

## Dimensionality reduction

First, we will use non-linear dimensionality reduction methods to project
cells from a high-dimensional (40) down to a low-dimensional (2) space. For this
the `r BiocStyle::Biocpkg("scater")` package provides the `runUMAP` and `runTSNE`
function. To ensure reproducibility, we will need to set a seed; 
however different seeds and different parameter settings (e.g. the `perplexity`)
parameter in the `runTSNE` function need to be tested to avoid interpreting
visualization artefacts. For dimensionality reduction, we will use all channels
that show biological variation across the dataset. However, marker selection 
can be performed with different biological questions in mind.

```{r dimred, message=FALSE}
library(scater)

set.seed(220225)
spe <- runUMAP(spe, subset_row = rowData(spe)$use_channel, exprs_values = "exprs") 
spe <- runTSNE(spe, subset_row = rowData(spe)$use_channel, exprs_values = "exprs") 
```

After dimensionality reduction, the low-dimensional embeddings are stored in the 
`reducedDim` slot.

```{r show-dimred-slot}
reducedDims(spe)
head(reducedDim(spe, "UMAP"))
```

Visualization of the low-dimensional embedding facilitates assessment of 
potential "batch effects". The `dittoDimPlot`
function allows flexible visualization. It returns `ggplot` objects which
can be further modified.

First, we will visualize single-cell metadata such as the patient ID and
the cancer type.

```{r visualizing-dimred-1, message=FALSE}
library(patchwork)
library(dittoSeq)
library(viridis)

# visualize patient id 
p1 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "UMAP", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$patient_id) +
    ggtitle("Patient ID on UMAP")
p2 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "TSNE", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$patient_id) +
    ggtitle("Patient ID on TSNE")

# visualize indication
p3 <- dittoDimPlot(spe, var = "indication", reduction.use = "UMAP", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$indication) +
    ggtitle("Indication on UMAP")
p4 <- dittoDimPlot(spe, var = "indication", reduction.use = "TSNE", size = 0.2) + 
    scale_color_manual(values = metadata(spe)$color_vectors$indication) +
    ggtitle("Indication on TSNE")

(p1 + p2) / (p3 + p4)
```

Next, we can visualize marker expression on the UMAP embedding.

```{r, visualizing-dimred-2, message=FALSE}
# visualize marker expression
p1 <- dittoDimPlot(spe, var = "Ecad", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "Ecad") +
    ggtitle("E-Cadherin expression on UMAP")
p2 <- dittoDimPlot(spe, var = "CD45RO", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "CD45RO") +
    ggtitle("CD45RO expression on UMAP")
p3 <- dittoDimPlot(spe, var = "CD20", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "CD20") +
    ggtitle("CD20 expression on UMAP")
p4 <- dittoDimPlot(spe, var = "CD3", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "CD3") +
    ggtitle("CD3 expression on UMAP")

(p1 + p2) / (p3 + p4)
```

We observe a strong separation of tumor cells (Ecad+ cells) between the
patients. Here, each patient was diagnosed with a different tumor type. The
separation of tumor cells could be of biological origin since tumor cells tend
to display differences in expression between patients and cancer types and/or of
technical origin: the panel only contains a single tumor marker (E-Cadherin) and
therefore slight technical differences in staining causes visible separation
between cells of different patients. Nevertheless, the immune compartment
(CD45RO+ cells) mix between patients and we can rule out systematic staining
differences between patients.

## Visualizing marker expression 

This section focuses on visualizing the expression of all markers and 
highlighting variation between cells, images and patients.

### Per cell

First, we will visualize single-cell marker expression in form of a
heatmap. Here, we sub-sample the dataset to 2000 cells for visualization
purposes and overlay the cancer type from which the cells were extracted.

```{r segmentation-heatmap, message=FALSE, fig.height=7}
cur_cells <- sample(seq_len(ncol(spe)), 2000)

dittoHeatmap(spe[,cur_cells], genes = rownames(spe)[rowData(spe)$use_channel],
             assay = "exprs", cluster_cols = TRUE, scale = "none",
             heatmap.colors = viridis(100), annot.by = "indication",
             annotation_colors = list(indication = metadata(spe)$color_vectors$indication))
```

We can differentiate between epithelial cells (Ecad+) and immune cells (CD45RO).
Some of the markers are specifically detected (e.g., Ki67, CD20, Ecad) and
others are more broadly detected (e.g. HLADR, B2M, CD4).

### Per image

It can be beneficial to visualize the mean marker expression per image
to identify images with outlying marker expression. This check does not
indicate image quality _per se_ but can highlight biological differences. Here,
we will use the `aggregateAcrossCells` function of the 
`r BiocStyle::Biocpkg("scuttle")` package to compute the mean expression per 
image. For visualization purposes, we again `asinh` transform the mean expression
values.

```{r mean-expression-per-image, message=FALSE, fig.height=7}
library(scuttle)

image_mean <- aggregateAcrossCells(spe, 
                                   ids = spe$sample_id, 
                                   statistics="mean",
                                   use.assay.type = "counts")
assay(image_mean, "exprs") <- asinh(counts(image_mean))

dittoHeatmap(image_mean, genes = rownames(spe)[rowData(spe)$use_channel],
             assay = "exprs", cluster_cols = TRUE, scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("indication", "patient_id", "ROI"),
             annotation_colors = list(indication = metadata(spe)$color_vectors$indication,
                                      patient_id = metadata(spe)$color_vectors$patient_id,
                                      ROI = metadata(spe)$color_vectors$ROI),
             show_colnames = TRUE)
```

### Per patient

The data presented here originate from samples from different hospitals with
potential differences in pre-processing and each sample stained individually.
These (and other) technical aspects can induce staining differences between
samples or batches of samples. In addition, patients were diagnosed with different 
cancer types. We will use ridgeline visualizations to check
differences in staining patterns and biological differences in expression:

```{r ridges, message=FALSE, fig.width=7, fig.height=25}
multi_dittoPlot(spe, vars = rownames(spe)[rowData(spe)$use_channel],
               group.by = "patient_id", plots = "ridgeplot", 
               assay = "exprs", 
               color.panel = metadata(spe)$color_vectors$patient_id)
```

We observe variations in the distributions of marker expression across patients.
These variations may arise partly from different abundances of cells in
different images (e.g. Patient3 may have higher numbers of CD11c+ and PD1+
cells) as well as staining differences between samples. While most of the
selected markers are specifically expressed in immune cell subtypes, we can see
that E-Cadherin (a marker for epithelial (tumor) cells) shows similar expression
across all patients.

## Image area covered by cells

A quality indicator for region selection is the image area covered by cells (or
biological tissue). This metric identifies regions of interest (ROIs) where
little cells are present, possibly hinting at incorrect selection of the ROI. We
can compute the percentage of covered image area using the metadata contained in
the `SpatialExperiment` object:

```{r cell-density, message = FALSE}
library(dplyr)

colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_area = sum(area),
           no_pixels = mean(width_px) * mean(height_px)) %>%
    mutate(covered_area = cell_area / no_pixels) %>%
    ggplot() +
        geom_point(aes(reorder(sample_id,covered_area), covered_area)) + 
        theme_minimal(base_size = 15) +
        ylim(c(0, 1)) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("% covered area") + xlab("")
```

## Cell size

Next, we observe the distributions of cell size across the individual images.
Differences in cell size distributions can indicate segmentation biases due to
differences in cell density or can indicate biological differences due to cell
type compositions (tumor cells tend to be larger than immune cells). 

```{r cell-size, message=FALSE}
colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    ggplot() +
        geom_boxplot(aes(sample_id, area)) +
        theme_minimal(base_size = 15) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cell area") + xlab("")

summary(spe$area)
```

The median cell size is `r median(spe$area)` pixels with a median major axis
length of `r round(median(spe$axis_major_length), digits = 1)`. The largest cell
has an area of `r max(spe$area)` pixels which relates to a diameter of 
`r round(sqrt(max(spe$area)), digits = 1)` pixels assuming a circular shape.
Overall, the distribution of cell sizes is similar across images with images from
`Patient4_005` and `Patient4_007` showing a reduced average cell size. These
images contain fewer tumor cells which can explain the smaller average cell size.

We detect very small cells in the dataset and will remove them.
The chosen threshold is arbitrary and needs to be adjusted per dataset.

```{r remove-small-cells}
sum(spe$area < 5)
spe <- spe[,spe$area >= 5]
```

## Spatial visualization

Here, we introduce the
[plotSpatial](https://bodenmillergroup.github.io/imcRtools/reference/plotSpatial.html)
function of the
[imcRtools](https://www.bioconductor.org/packages/release/bioc/html/imcRtools.html)
package to visualize the cells' centroids and cell-cell interactions as spatial
graphs.

In the following example, we select images of one patient for visualization
purposes. Here, each dot (node) represents a cell and edges are drawn between
cells in close physical proximity as detected by `steinbock`. Nodes are variably
colored based on the cells' expression level of E-cadherin and their size can be
adjusted (e.g., based on measured cell area).

```{r spatial-viz-1, message=FALSE, fig.width=12, fig.height=12}
library(ggplot2)
library(viridis)

# steinbock interaction graph 
plotSpatial(spe[,grepl("Patient3", spe$sample_id)], 
            node_color_by = "Ecad", 
            assay_type = "exprs",
            img_id = "sample_id", 
            draw_edges = TRUE, 
            colPairName = "neighborhood", 
            nodes_first = FALSE, 
            node_size_by = "area", 
            directed = FALSE,
            edge_color_fix = "grey") + 
    scale_size_continuous(range = c(0.1, 2)) +
    ggtitle("E-cadherin expression")
```

# Image visualization

The following section highlights the use of the `r BiocStyle::Biocpkg("cytomapper")` package to visualize multichannel images and
segmentation masks. For visualization purposes we select 3 images from the 
dataset:

```{r select-images, message=FALSE}
# Sample images
set.seed(220517)
cur_id <- sample(unique(spe$sample_id), 3)
cur_images <- images[names(images) %in% cur_id]
cur_masks <- masks[names(masks) %in% cur_id]
```

## Pixel visualization 

The following section gives examples for visualizing individual channels or 
multiple channels as pseudo-color composite images. For this the `cytomapper`
package exports the `plotPixels` function which expects a `CytoImageList` object
storing one or multiple multi-channel images. In the simplest use case, a 
single channel can be visualized as follows:

```{r single-channel}
plotPixels(cur_images, 
           colour_by = "Ecad",
           bcg = list(Ecad = c(0, 5, 1)))
```

The plot above shows the tissue expression of the epithelial tumor marker
E-cadherin on the 3 selected images. The `bcg` parameter (default `c(0, 1, 1)`)
stands for "background", "contrast", "gamma" and controls these attributes of
the image. This parameter takes a named list where each entry specifies these
attributes per channel. The first value of the numeric vector will be added to
the pixel intensities (background); pixel intensities will be multiplied by the
second entry of the vector (contrast); pixel intensities will be exponentiated
by the third entry of the vector (gamma). In most cases, it is sufficient to
adjust the second (contrast) entry of the vector.

The following example highlights the visualization of 6 markers (maximum allowed
number of markers) at once per image. The markers indicate the spatial
distribution of tumor cells (E-cadherin), T cells (CD3), B cells (CD20), CD8+ T
cells (CD8a), plasma cells (CD38) and proliferating cells (Ki67).

```{r 6-channel}
plotPixels(cur_images, 
           colour_by = c("Ecad", "CD3", "CD20", "CD8a", "CD38", "Ki67"),
           bcg = list(Ecad = c(0, 5, 1),
                      CD3 = c(0, 5, 1),
                      CD20 = c(0, 5, 1),
                      CD8a = c(0, 5, 1),
                      CD38 = c(0, 8, 1),
                      Ki67 = c(0, 5, 1)))
```

### Adjusting colors

The default colors for visualization are chosen by the additive RGB (red, green,
blue) color model. For six markers the default colors are: red, green, blue,
cyan (green + blue), magenta (red + blue), yellow (green + red). These colors
are the easiest to distinguish by eye. However, you can select other colors for
each channel by setting the `colour` parameter:

```{r setting-colors}
plotPixels(cur_images, 
           colour_by = c("Ecad", "CD3", "CD20"),
           bcg = list(Ecad = c(0, 5, 1),
                      CD3 = c(0, 5, 1),
                      CD20 = c(0, 5, 1)),
           colour = list(Ecad = c("black", "burlywood1"),
                         CD3 = c("black", "cyan2"),
                         CD20 = c("black", "firebrick1")))
```

The `colour` parameter takes a named list in which each entry specifies the
colors from which a color gradient is constructed via `colorRampPalette`. These
are usually vectors of length 2 in which the first entry is `"black"` and the
second entry specifies the color of choice. Although not recommended, you can
also specify more than two colors to generate a more complex color gradient.

### Image normalization

As an alternative to setting the `bcg` parameter, images can first be
normalized. Normalization here means to scale the pixel intensities per channel
between 0 and 1 (or a range specified by the `ft` parameter in the `normalize`
function). By default, the `normalize` function scales pixel intensities across
**all** images contained in the `CytoImageList` object (`separateImages = FALSE`).
Each individual channel is scaled independently (`separateChannels = TRUE`).

After 0-1 normalization, maximum pixel intensities can be clipped to enhance the
contrast of the image (setting the `inputRange` parameter). In the following
example, the clipping to 0 and 0.2 is the same as multiplying the pixel
intensities by a factor of 5.

```{r default-normalization}
# 0 - 1 channel scaling across all images
norm_images <- normalize(cur_images)

# Clip channel at 0.2
norm_images <- normalize(norm_images, inputRange = c(0, 0.2))

plotPixels(norm_images, 
           colour_by = c("Ecad", "CD3", "CD20", "CD8a", "CD38", "Ki67"))
```

The default setting of scaling pixel intensities across all images ensures 
comparable intensity levels across images. Pixel intensities can also be 
scaled **per image** therefore correcting for staining/expression differences
between images:

```{r individual-normalization}
# 0 - 1 channel scaling per image
norm_images <- normalize(cur_images, separateImages = TRUE)

# Clip channel at 0.2
norm_images <- normalize(norm_images, inputRange = c(0, 0.2))

plotPixels(norm_images, 
           colour_by = c("Ecad", "CD3", "CD20", "CD8a", "CD38", "Ki67"))
```

As we can see, the marker Ki67 appears brighter on image 2 and 3 in comparison
to scaling the channel across all images.

Finally, the `normalize` function also accepts a named list input for the
`inputRange` argument. In this list, the clipping range per channel can be set
individually:

```{r setting-inputRange}
# 0 - 1 channel scaling per image
norm_images <- normalize(cur_images, 
                         separateImages = TRUE,
                         inputRange = list(Ecad = c(0, 50), 
                                           CD3 = c(0, 30),
                                           CD20 = c(0, 40),
                                           CD8a = c(0, 50),
                                           CD38 = c(0, 10),
                                           Ki67 = c(0, 70)))

plotPixels(norm_images, 
           colour_by = c("Ecad", "CD3", "CD20", "CD8a", "CD38", "Ki67"))
```

## Segmentation mask visualization

The `cytomapper` package provides the `plotCells` function to visualize the
aggregated pixel intensities per cell on segmnetation masks. In the current
dataset pixel intensities were aggregated by computing the mean pixel intensity
per cell and per channel. The `plotCells` function accepts the `exprs_values`
argument (default `counts`) that allows selecting the assay which stores the
expression values that should be visualized.

In the following example, we visualize the asinh-transformed mean pixel
intensities of the epithelial marker E-cadherin on segmentation masks.

```{r Ecad-expression}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "Ecad",
          exprs_values = "exprs")
```

## Segmentation quality control {#seg-quality}

We can now use the `plotPixels` function to outline segmented cells on image
composites to observe segmentation accuracy. Without having ground-truth data
readily available, a common approach to segmentation quality control is to
overlay segmentation masks on composite images displaying channels that were
used for segmentation.

Here, we select 3 random images and perform image- and channel-wise
normalization (channels are first min-max normalized and scaled to a range of
0-1 before clipping the maximum intensity to 0.2).

```{r overlay-masks, message=FALSE}
library(cytomapper)
set.seed(20220118)
img_ids <- sample(seq_len(length(images)), 3)

# Normalize and clip images
cur_images <- images[img_ids]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD163", "CD20", "CD3", "Ecad", "DNA1"),
           colour = list(CD163 = c("black", "yellow"),
                         CD20 = c("black", "red"),
                         CD3 = c("black", "green"),
                         Ecad = c("black", "cyan"),
                         DNA1 = c("black", "blue")),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
```

We can see that nuclei are centered within the segmentation masks and all cell
types are correctly segmented . A common challenge here is to segment large (e.g.
epithelial cells - in cyan) _versus_ small (e.g. B cells - in red). However, the
segmentation approach here appears to correctly segment cells across different
sizes.

## Save objects

Finally, the generated data objects can be saved for further downstream 
processing and analysis.

```{r save-objects-read-data}
saveRDS(spe, "data/spe.rds")
saveRDS(images, "data/images.rds")
saveRDS(masks, "data/masks.rds")
```

## Obtaining public IMC datasets

The [imcdatasets](https://bioconductor.org/packages/release/data/experiment/html/imcdatasets.html) 
R/Bioconductor package provides publicly available IMC datasets in form 
of `SingleCellExperiment` (single-cell data) and `CytoImageList` (images and
segmentation masks) objects.

```{r imcdatasets, eval=FALSE, message=FALSE}
library(imcdatasets)

listDatasets()

sce <- Damond_2019_Pancreas(data_type = "sce")
images <- Damond_2019_Pancreas(data_type = "images")
masks <- Damond_2019_Pancreas(data_type = "masks")
```

## Session info

```{r}
sessionInfo()
```
