---
title: "Data download"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For session 2, we will download IMC example data to highlight the use of the
`steinbock` framework for image processing.

The example data were acquired as part of the **I**ntegrated i**MMU**noprofiling
of large adaptive **CAN**cer patient cohorts projects
([immucan.eu](https://immucan.eu/)). The raw data of 4 patients can be accessed
online at [zenodo.org/record/5949116](https://zenodo.org/record/5949116). Here,
we will also download a panel file which matches the raw data as well as
metadata regarding the samples.

```{r download-raw, message=FALSE, warning=FALSE}
options(timeout=10000)
if (!dir.exists("data/steinbock")) dir.create("data/steinbock")
if (!dir.exists("data/steinbock/raw")) dir.create("data/steinbock/raw")

download.file("https://zenodo.org/record/7412972/files/panel.csv", 
              "data/steinbock/panel.csv")
download.file("https://zenodo.org/record/5949116/files/Patient1.zip",
              "data/steinbock/raw/Patient1.zip")
download.file("https://zenodo.org/record/5949116/files/Patient2.zip",
              "data/steinbock/raw/Patient2.zip")
download.file("https://zenodo.org/record/5949116/files/Patient3.zip",
              "data/steinbock/raw/Patient3.zip")
download.file("https://zenodo.org/record/5949116/files/Patient4.zip",
              "data/steinbock/raw/Patient4.zip")
download.file("https://zenodo.org/record/5949116/files/sample_metadata.xlsx", 
         destfile = "data/steinbock/sample_metadata.xlsx")
```

If you can't use `steinbock` for image processing we also provide the resulting
files for download:

```{r download-steinbock}
# download intensities
url <- "https://zenodo.org/record/7412972/files/intensities.zip"
destfile <- "data/steinbock/intensities.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock/", overwrite=TRUE)
unlink(destfile)

# download regionprops
url <- "https://zenodo.org/record/7412972/files/regionprops.zip"
destfile <- "data/steinbock/regionprops.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock/", overwrite=TRUE)
unlink(destfile)

# download neighbors
url <- "https://zenodo.org/record/7412972/files/neighbors.zip"
destfile <- "data/steinbock/neighbors.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock/", overwrite=TRUE)
unlink(destfile)

# download images
url <- "https://zenodo.org/record/7412972/files/img.zip"
destfile <- "data/img.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock/", overwrite=TRUE)
unlink(destfile)

# download masks
url <- "https://zenodo.org/record/7412972/files/masks_deepcell.zip"
destfile <- "data/steinbock/masks_deepcell.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock/", overwrite=TRUE)
unlink(destfile)

# download individual files
download.file("https://zenodo.org/record/7412972/files/images.csv", 
              "data/steinbock/images.csv")
```

### Files for spillover matrix estimation

To highlight the estimation and correction of channel-spillover as described by
[https://www.cell.com/cell-systems/fulltext/S2405-4712(18)30063-2](https://www.cell.com/cell-systems/fulltext/S2405-4712(18)30063-2), we can access an example spillover-acquisition from
[zenodo.org/record/5949116](https://zenodo.org/record/5949116)

```{r download-spillover-data}
download.file("https://zenodo.org/record/5949116/files/compensation.zip",
              "data/compensation.zip")
unzip("data/compensation.zip", exdir="data", overwrite=TRUE)
unlink("data/compensation.zip")
```

### Gated cells

In Session 4, we present a cell type classification approach
that relies on previously gated cells. This ground truth data is available
online at [zenodo.org/record/7432486](https://zenodo.org/record/7432486) and
will be downloaded here for later use:

```{r download-gated-cells}
download.file("https://zenodo.org/record/7432486/files/gated_cells.zip",
              "data/gated_cells.zip")
unzip("data/gated_cells.zip", exdir="data", overwrite=TRUE)
unlink("data/gated_cells.zip")

unlink("data/__MACOSX", recursive = TRUE)
```